Was Devon Allen Unjustly Disqualified at the 2022 World Track and Field Championships?

### 2023-03-21 Manuscript/manuscript.tex
Small changes

One of the most interesting things found was that the reaction times from the USA
athletes in the World Championship 110 meter hurdles, was that they were significantly
faster than at the USA Track Championships.  From June 23-26 2022, USA held it's Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all of the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teamates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  This table below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. "USA" denotes
the USA Track and Field Championships, "World" denotes the World Track and Field
Championships, "N/A" was used for any athlete that did not compete and thus
did not register a reaction time.  All numbers listed in the table are the reaction
time for the respective athletes in seconds.

@@

\section{GLMM Gamma Model}
When determining the most effective model to use, several examples were
explored across the 4 different data sets: a linear model without a year effect,
linear model with year effect, a gamma model without a year effect, a gamma model
with a year effect, a gamma model without a year effect and link set to log instead
of the default which is Gaussian.  The R package lme4 was very helpful with this
as it supported the Generalized Linear Mixed Model (GLMM) which allowed both
linear and gamma mixed effects models to be built \citep{Rpkg:lme4}. A GLMM is a type of model
that supports inputs that are from different distributions.  The model that we use is for the
Gamma distribution for the fixed effects part of the model and a random effects part.  In
the model used in the paper, the fixed effect comes in the form of an intercept, and then
there is a random effect which in this case we will call the venue effect.

@@

@@

The equipment used to measure reaction time cannot be guaranteed
to be the same from national to international competition, but the time span between
meets has been reduced considerably.  Repeating the data collection steps and
again performing a clustered analysis for a Wilcoxon ranked sum test
results in two new z scores and p values.  The "ds" method produced a z score of
3.6069 and p value of 0.000155.  This result is highly significant and shows that
the times at the national stage of competition were significantly higher than
the reaction times at the World Championships.

@@

A simulation was performed on the gamma model with the year effect and link set
to "log". At a size of n=12,000,000, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:

@@

>>>>

One of the most interesting things found was that the reaction times from the USA
athletes in the World Championship 110 meter hurdles, was that they were significantly
faster than at the USA Track Championships.  From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all of the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teamates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  This table below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. "USA" denotes
the USA Track and Field Championships, "World" denotes the World Track and Field
Championships, "N/A" was used for any athlete that did not compete and thus
did not register a reaction time.  All numbers listed in the table are the reaction
time for the respective athletes in seconds.

@@

\section{GLMM Gamma Model}
When determining the most effective model to use, several examples were
explored across the 4 different data sets: a linear model without a year effect,
linear model with year effect, a gamma model without a year effect, a gamma model
with a year effect, a gamma model without a year effect and link set to log instead
of the default which is Gaussian.  The R package lme4 was very helpful with this
as it supported the Generalized Linear Mixed Model (GLMM) which allowed both
linear models with various distributions to be modeled\citep{Rpkg:lme4}. A GLMM is a type of model
that supports inputs that are from different distributions.  The model that we use is for the
Gamma distribution for the fixed effects part of the model and a random effects part.  In
the model used in the paper, the fixed effect comes in the form of an intercept, and then
there is a random effect which in this case we will call the venue effect.  The
venue effect is responsible for the variation between years and tries to explain
why some years have athletes with faster reaction times than others.

@@

The effect of adding link = "log" is that

@@

The equipment used to measure reaction time cannot be guaranteed
to be the same from national to international competition, but the time span between
meets has been reduced considerably.  Repeating the data collection steps and
again performing a clustered analysis for a Wilcoxon ranked sum test
results in two new z scores and p values.  The "ds" method produced a z score of
3.6069 and p value of 0.000155.  This result is highly significant and shows that
the times at the national stage of competition were significantly higher than
the reaction times at the World Championships even for an alpha level of 0.001.

@@

To try and approximate the distribution of the reaction times, R code was written
to try and estimate the probabilites of observing a reaction time as extreme as
0.1 seconds.At a size of n=12,000,000, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:

@@

It is interesting that the probability of observing a reaction time below 0.1
seconds increases drastically for the finals data and increases by a smaller margin
for the pooled data set after 2022 is included in the data set.  By including 2022
in the data, the chances of observing an extreme time are increased, showing that
2022 is a high outlier for reaction times compared to other years.
